Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models

### 2014-11-24 NewDesignDoc/DesignDocGLMM.tex
Editing, rearranging design doc

Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data. Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved random effects. Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects. Let $\theta=(\beta ,  \nu)'$ be a vector containing all unknown parameters. Then the data $y$ are distributed conditionally on the random effects according to $f_\theta(y|u)$ and the random effects are distributed according to $f_\theta(u)$. Although $f_\theta(u)$ does not actually depend on $\beta$ and $f_\theta(y|u)$ does not depend on $\nu$, we write both densities with $\theta$ to keep notation simple in future equations.

Since $u$ is unobservable, the log likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\log \int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Rather than evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is a mixture distribution specified in section \ref{sec:ftwiddle}.
Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
\end{align}
When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
&\nabla^2 l_m(\theta)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}.
\end{align}
Details for these equations can be found in section \ref{sec:calcs}. Calculation for the gradient and hessian when $\tilde{f}$ is independent of $\theta$ can be found in section \ref{sec:calcsindep}.

@@

%Let $y=(y_1, \ldots, y_n)^T$ be a vector of observed data. Let $u=(u_1,\ldots,u_q)^T$ be a vector of unobserved random effects. Let $\theta$ denote a vector of unknown parameters.  The data are distributed conditionally on the random effects according to $f_\theta(y|u)$.

\section{Notation}

 Let $X$ be an $n \times p$ design matrix for the fixed effects.  Let $U$ be an unobservable normal random vector with length $q$, mean $0$ and variance matrix $D$. Let $Z$ be a $n \times q$ model matrix for the random effects.

There are $T$ distinct variance components  $\nu_t$, $t = 1,$ $\ldots,$ $T$.  For the first version of the package, $D$ is assumed diagonal. Let $E_t$ be a diagonal matrix with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$. (Learn more about $E_t$ in section \ref{sec:getEk}.) Then $D= \sum_{t=1}^T \nu_t E_t$.  Later versions of this package will allow  more general covariance structures. With other structures, $D$ will be constructed by $D= \sum_{t=1}^T h_t (\nu_t) \; E_t$, but $E_t$ will no longer be diagonal and $h_t$ will be some function. We will have a D for distance (such as in the car theft data) and for autoregressive 1.

 Denote the density of $U$ by $f_\theta(u)$  with $\theta= (\beta,\nu_{1},\ldots,\nu_T)$. Although $f_\theta(u)$ does not actually depend on $\beta$, we write the density like this to keep notation simple. Since $U$ is normally distributed, we can write its log density as
\begin{align}
\log f_\theta(u) = -\dfrac{1}{2} \log |D| -\dfrac{1}{2} U' D^{-1} U
\end{align}
where $|D|$ denotes the determinant of $D$.  More details on the distribution of the random effects can be found in section \ref{sec:distRand}.

Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
\begin{align}
g(\mu) = X \beta + Z U
\end{align}
The choice of the link function is related to the distribution of the data, $\log f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$. More on these two families can be found in section \ref{sec:fam}. For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
\begin{align}
Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
\end{align}

\section{Model fitting function} 
This will be the main function the user will use. The users will need to specify the response and the predictors using the R formula mini-language as interpreted by model.matrix. They'll need to specify the  family (either binomial or Poisson, though  really any exponential family would work and I could add more later). The user will specify the random effects in the same way as for the R function \texttt{reaster} in the R package
 \texttt{aster} \citep{aster-package}. That is, random effects will be expressed using the R formula mini-language. Thus, a sample command with fixed predictors $x_1$ and $x_2$ and with random effects $school$ and $classroom$ (in data set as categorical variables ) would look like
\begin{verbatim}
glmm(y ~ x1+ x2, list(~0+school,~0+classroom),  family.glmm="binomial.glmm", 
data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2),
debug=FALSE )
 \end{verbatim} 
Section \ref{sec:fam} contains more information on the family.

@@

It's possible that the user could want some variance components to be set equal. For example, in the Coull and Agresti flu problem, there are 4 years and random effects for each year. The authors want the within-year variance components to be equal.  There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since year 4's virus was a repeat). Suppose year is a categorical variable with four levels, and year1 through year4 are dummy variables. Thus, the call could contain these arguments
\begin{verbatim}
glmm(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeet,~0+year3,~0+year4),
 varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeet"),
data=flu,family.glmm=bernoulli.glmm)
\end{verbatim}

Thus model.matrix will make Z into a list of 6 design matrices. Since we have 3 distinct variance components, we want 3 design matrices. What we do is we take the design matrices that share a variance component and then cbind them together. Thus we'll have 3 model matrices in our Z list, one for each random effect. We will want to put them in order (1,2,3 according to varcomps.equal) so that the names for each matrix line up ("subject","year","repeet").  Cbinding the design matrices that share a variance component will not effect the way that we compute $\eta$.  Also, the variance estimates should come out in this same order as well.

  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.  The process of finding these estimates is detailed in section \ref{sec:pql}. The PQL estimates  will help determine the means and variances for the normal distributions from which random effects will be generated. After  sampling random effects from a standard normal, they will be scaled and centered to have the desired mean and variance. More information on generated the random effects is in section \ref{sec:genRand}. Next, I'll implement trust on the objective function (details on the objective function are in section \ref{sec:objfun}. Finally, the function will return parameter estimates, the log likelihood evaluated at those estimates, the gradient vector of the log likelihood, and the Hessian of the log likelihood at those estimates.

%%Families%%
\section{Families} \label{sec:fam}
This function will be hidden from the user.  These functions (along with the distribution of random effects) are necessary to approximate the log likelihood.

I'm going to have an S3 class called ``glmm.family'' with a few options (for now, binomial and poisson). Each family  function will output a list including the family name (a character string such as ``binomial''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.

The user will provide the family in the model-fitting function. They can either enter the character string (``bernoulli.glmm''), the function (bernoulli.glmm()), or the result of invoking it.  The following code for using the input to determine the family is adapted from glm.

@@

For example, one of the family functions will look like this:
\begin{verbatim}
poisson.glmm<-function()
{
	family.glmm<- "poisson.glmm"
	c <- function(eta) exp(eta)
	cp <- function(eta) exp(eta)
	cpp<-function(eta) exp(eta)
	out<-list(family.glmm=family.glmm, c=c,cp=cp,cpp=cpp)
	class(out)<-"glmm.family"
	return(out)
}
\end{verbatim}

Then, to use these functions in order to calculate $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$, I can just call 
\begin{verbatim}
family.glmm$c(args)
family.glmm$cp(args)
family.glmm$cpp(args)
\end{verbatim}

For  Bernoulli, we calculate these values (c, cp, and cpp) with careful computer arithmetic as follows. We also use the \texttt{log1p} function in R for $c(\eta_i)$. 
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}
We use the \texttt{log1p} function in R for $c(\eta_i)$ for the Bernoulli family. For poisson, these values (c,cp, and cpp) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

@@

In the R function \texttt{glm} binomial, the user can choose the link. The canonical link must be used in the \texttt{glmm} package so that we have an exponential family. I'm going to include the link in the value of \texttt{family.glmm} just in case the user doesn't know it already.

Also, I'll have a function to check that the data are valid given the family type. So I'll make sure that if \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should be 0's or 1's. If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative integers.  If that's not true, the check returns an error message.

\subsection{Redone in C}
When I redo this in C, I will have a separate function for \texttt{cum} ,\texttt{cp}, and \texttt{cpp}. The inputs will be \texttt{eta} (an array of doubles), $\texttt{neta}$ (the length of \texttt{eta}), the type (to denote the family), and an array of doubles to contain the result. These will be passed in as pointers. The functions will calculate  the cumulant function or one of its first two derivatives. Each function will contain a switch statement for the glmm family. The calculations for each of these functions have  been shown earlier in this section. This function will be type void: rather than returning the cumulant or its derivatives, the pointers will be changed to contain the results. This function will be invoked by \texttt{el}, descriped in section \ref{sec:el}.

@@

\subsection{Redone in C}
The C function to calculate the value of the log data density and its two derivatives will be called by reference. The following pointers will be passed in: double \texttt{Y}, double \texttt{X}, int \texttt{nrowX}, int \texttt{ncolX}, double \texttt{eta}, int \texttt{family}, double \texttt{elval}, double \texttt{elgradient}, double \texttt{elhessian}. The pointers  \texttt{elval}, \texttt{elgradient} and \texttt{elhessian} will be zeros before \texttt{el.C} is invoked. Invoking \texttt{el.C} will then place the calculatedvalue of the log data density and two derivatives into \texttt{elval}, \texttt{elgradient} and \texttt{elhessian}.\\

The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, texttt{cpp3.C} to calculate the hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.

%%getEk
\section{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
Since $D$ is a function of $E_t$, we need to find a way to calculate $E_t$. This depends on whether $D$ is diagonal or not.

\subsection{D diagonal}
When $D$ is diagonal, recall $E_t$ are diagonal with either 1 or 0 on the diagonal (1 if that random effect has $\nu_t$ as a variance component and 0 otherwise). Let $q_t$ be the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$. When D is diagonal, then $\sum_{t=1}^T E_t = I_q$, the identity matrix with $q$ rows. Finally, recall that $D=\sum_{t=1}^T \nu_t E_t$.

At this point,  \texttt{mod.mcml\$z} is a list with $T$ design matrices (one for each distinct variance component).  We need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$ will have that many rows (and columns). Thus, each $E_t$ will have q rows and columns as well.  Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.

\subsection{D in general (not necessarily diagonal)}
In progress.

\section{Generating random effects (genRand)}\label{sec:genRand}

This will be hidden from the user. It will be called by the model fitting function.

To approximate the log likelihood, we need simulated random effects.  PQL will give us estimates for $\beta$, $\sigma$ and  $s$. Let $\beta^*$, $s^*$ and $\sigma^*$ be the vectors of PQL estimates. Then 
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*.
\end{align}
Also, we can ``unstandardize'' our random effects this way:
\begin{align}
u^*&=A^*s^*
\end{align}

I'll be generating random effects $u$ from these distributions:
\begin{enumerate}
\item $N(0,I)$
\item $N(u^*,D^*)$
\item $N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$
\end{enumerate}
I'll generate them in these proportions: $p_1,p_2,p_3$ where $p_1+p_2+p_3=1$.

The first distribution is centered at 0, just where we expect the random effects to truly be centered. The random effects will be generated with variance 1, but will be scaled in the trust objective function so that they have variance $\nu$, where $\nu$ is given by the iteration of trust.   The second distribution is centered at the PQL best guess of the random effect values and  has the PQL guess of the variance. The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood hessian.  The idea of this last distribution is to generate random effects from a distribution whose first two derivatives are equal to those of the target distribution $f_\theta(u,y)$.

Let $u_k$ be a vector with length q (the number of random effects in the model). Suppose we want to generate random effects $u_k,k=1,...,m$ from a generic distribution $N(\mu,\Sigma)$.  We can use eigendecomposition for $\Sigma$ to get orthogonal matrix $O$ (containing the eigenvectors) and diagonal matrix $\Lambda$ with diagonal entries being $\Sigma$'s eigenvalues $\lambda$. Eigendecompositions take a little bit more time than Cholesky decompositions, but are more stable. Then
\begin{align}
\Sigma^{1/2}= O \Lambda^{1/2} O'
\end{align}
Finding $\Lambda^{1/2}$ is as easy as taking the root of the diagonal entries.  We know that $\Sigma^{1/2}$ is correct because
\begin{align}
\Sigma^{1/2} \Sigma^{1/2} &= O \Lambda^{1/2} O'O \Lambda^{1/2} O'\\
&= O \Lambda^{1/2}\Lambda^{1/2} O'\\
&=O \Lambda O' \\
&= \Sigma.
\end{align}

For each k, draw a sample of size q from the standard normal distribution. Call this sample $\breve{u}_k$. Then we shift and scale it:
\begin{align}
u_k = \breve{u}_k \Sigma^{1/2} + \mu.
\end{align}
This is the process of drawing our sample from $N(\mu,\Sigma^{1/2}\Sigma^{1/2})$ for any mean vector $\mu$ and variance matrix $\Sigma$.

%%distribution of the random effects%%
\section{Distribution of the random effects }
This will be hidden from the user.  This section provides equations for evaluating the log of a normal distribution with a specified mean and variance. When necessary to emphasize and clarify the choice of mean $\mu$ and variance $\Sigma$ , I use the notation $ f(u|\mu,\Sigma)$ rather than $f_\theta(u)$.

This section is useful for calculating $\tilde{f}(u_k)$.  This section also provides two derivatives in some cases, which is useful for evaluating $f_\theta(u_k)$ and two derivatives. These are needed for the calculation in \eqref{eq:MCLA}.

>>>>

Let $y=(y_1, \ldots, y_n)'$ be a vector of observed data. Let $u=(u_1,\ldots,u_q)'$ be a vector of unobserved random effects centered at 0 with variance matrix $D$. Let $\beta$ be a vector of $p$ fixed effect parameters and let $\nu$ be a vector of $T$ variance components for the random effects so that $D$ depends on $\nu$. Let $\theta=(\beta ,  \nu)'$ be a vector containing all unknown parameters. Then the data $y$ are distributed conditionally on the random effects according to $f_\theta(y|u)$ and the random effects are distributed according to $f_\theta(u)$. Although $f_\theta(u)$ does not actually depend on $\beta$ and $f_\theta(y|u)$ does not depend on $\nu$, we write both densities with $\theta$ to keep notation simple in future equations.

Since $u$ is unobservable, the log likelihood must be expressed by integrating out the random effects:
\begin{align}
l(\theta)=\log \int f_\theta(y|u) f_\theta(u) \; du
\end{align}
For most datasets, this integral is intractible. In these cases, performing even basic inference on the likelihood is not possible. Rather than evaluating the integral, \citet{geyer:thom:1992} suggest using a Monte Carlo approximation to the likelihood. Monte Carlo likelihood approximation (MCLA) uses an importance sampling distribution $\tilde{f}(u)$ to generate random effects $u_k, k=1, \ldots, m$ where $m$ is the Monte Carlo sample size.  MCLA theoretically works for any $\tilde{f}(u)$, but the  $\tilde{f}(u)$ chosen for this package is the mixture distribution specified in section \ref{sec:genRand}.

Then the Monte Carlo log likelihood approximation is
\begin{align}
l_{m}(\theta) &=\log \dfrac{1}{m} \sum_{k=1}^mf_\theta(y|u_k)  \dfrac{ f_\theta(u_k)   }{\tilde{f}(u_k)}\\
&= \log \dfrac{1}{m} \sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}.
\end{align}
When $\tilde{f}$ depends on $\theta$, the gradient vector of the MCLA with respect to $\theta$ is
\begin{align}
\nabla l_m(\theta)&= \dfrac{\sum_{k=1}^m    \left( \nabla \log f_\theta(u_k,y) - \nabla \log \tilde{f} (u_k)  \right) \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\sum_{k=1}^m \left( \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} \right) }, \label{eq:MCLAgradient}
\end{align}
and the Hessian matrix of the MCLA is
\begin{align}
&\nabla^2 l_m(\theta)= \dfrac{   \sum_{k=1}^m \left[ \nabla^2 \log f_\theta(y,u_k)  -
   \nabla^2 \log \tilde{f}(u_k)   \right]  \dfrac{ f_\theta(y,u_k)}{\tilde{f}(u_k)}  }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}\\
&+ \dfrac{   \sum_{k=1}^m \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) - \nabla l_m(\theta)   \right] \left[ \nabla \log f_\theta(y,u_k)  -
   \nabla \log \tilde{f}(u_k) -\nabla l_m(\theta)  \right]'  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}   }{\sum_{k=1}^m  \dfrac{ f_\theta(y,u_k)   }{\tilde{f}(u_k)}}.
\end{align}
Details for these equations can be found in section \ref{sec:calcs}. Calculation for the gradient and hessian when $\tilde{f}$ is independent of $\theta$ can be found in section \ref{sec:calcsindep}.

@@

\section{Model fitting function} 
This is be the primary function the user  uses. The user  specifies the response and the predictors using the R formula mini-language as interpreted by \texttt{model.matrix}. Let $n$ be the observed sample size and recall that $p$ is the number of fixed effects.  When the fixed effects are specified, the model matrix $X$  is created, where $X$ has dimensions $n \times p$.

The user  specifies the random effects in the same way as for the R function \texttt{reaster} in the R package
 \texttt{aster} \citep{aster-package}. That is, random effects are expressed using the R formula mini-language. Let $q_t$ be the number of random effects associated with variance component $\nu_t$. When the random effects are specified, a list of T model matrices are created and the $t$th model matrix $Z_t$ has dimensions $n \times q_t$.

The user also specifies the expontial family (details in section \ref{sec:fam}), the name of the data set, and the name of the variance components.

Thus, the following is a sample command with fixed predictors $x_1$ and $x_2$ and with random effects created by categorical variables $school$ and $classroom$:
\begin{verbatim}
glmm(y ~ x1+ x2, list(~0+school,~0+classroom),  family.glmm="bernoulli.glmm", 
data=schooldat,varcomps.names=c("school","classroom"),varcomps.equal=c(1,2),
debug=FALSE)
 \end{verbatim}

@@

It is possible that the user could want some variance components to be set equal. For example, \pcite{coull:agresti:2000} influenza dataset contains four years with random effects for each year. The authors want the within-year variance components to be equal.  There are also variance components for subject-specific intercepts and for the decreased susceptibility to illness in year 4 (since the strain of flu during year 4 was a repeat of a previous year). Suppose \texttt{year} is a categorical variable with four levels, and \texttt{year1} through \texttt{year4} are dummy variables. Thus, the call  contains these arguments:
\begin{verbatim}
glmm(y~year,list(~0+subject,~0+year1,~0+year2,~0+repeat,~0+year3,~0+year4),
 varcomps.equal=c(1,2,2,3,2,2), varcomps.names=c("subject","year","repeat"),
data=flu,family.glmm=bernoulli.glmm)
\end{verbatim}

As a result of this command, \texttt{model.matrix}  makes $Z$ into a list of 6 design matrices. Since we have 3 distinct variance components, we want 3 design matrices. We take the design matrices that share a variance component and \texttt{cbind} them together. The result is 3 model matrices in the $Z$ list, one for each variance component. We  want to put them in order (1,2,3 according to \texttt{varcomps.equal}) so that the names for each model matrix  ("subject","year","repeat") are correctly assigned.  Also, the variance estimates  come out in this same order as well.

  After interpreting the model the user has specified, the next step is to find penalized quasi-likelihood (PQL) estimates.  The process of finding these estimates is detailed in section \ref{sec:pql}. The PQL estimates parameterize the importance sampling distribution $\tilde{f}(u_k)$ which generate the random effects.  More information on generated the random effects is in section \ref{sec:genRand}.

Next, \texttt{trust} is implemented to maximize  the MCLA objective function (details on the objective function are in section \ref{sec:objfun}). Finally, the function  returns parameter estimates, the log likelihood evaluated at those estimates, the gradient vector of the log likelihood,  the Hessian of the log likelihood at those estimates, information from the \texttt{trust} optimization, and other information.

%%Families%%
\section{Families} \label{sec:fam}

Let $g$ be the canonical link function and $\mu$ be a vector of length $n$ such that
\begin{align}
g(\mu) = X \beta + Z U
\end{align}
The choice of the link function is related to the distribution of the data, $ f_\theta(y|u)$. If the data have a  Bernoulli distribution, the link is $\logit(\mu)$. If the data have a Poisson distribution, the link is $\log (\mu)$. Currently, \texttt{glmm} offers only these two choices for the family, but any exponential family could work and can be easily added later.

 For simplicity of future notation, let $\eta=g(\mu)=X \beta + Z U$. Let $c(\eta)$ denote the cumulant function such that the log of the data density can be written as
\begin{align}
Y' \eta - c(\eta) = \sum_i \left[ Y_i \eta_i - c(\eta_i)  \right]
\end{align}

The user is required to specify the family in the model-fitting function. Once the family is specified, many family-specific functions are called. They are contained in an S3 class called ``glmm.family''. Each family function outputs a list including the family name (a character string such as ``bernoulli.glmm''), a function that calculates the value of the cumulant function $c(\eta)$,  a function that calculates the cumulant's first derivative $c'(\eta)$ with the derivative taken with respect to $\eta$, and  a function that calculates the cumulant's second derivative $c''(\eta)$.

The users provide the family in the model-fitting function by either enter the character string (``bernoulli.glmm''), the function (\texttt{bernoulli.glmm()}), or the result of invoking the function.  The following code for using the input to determine the family is adapted from \texttt{glm}.

@@

With this, calculating $c(\eta_i), c'(\eta_i),$ and  $c''(\eta_i)$ is as simple as: 
\begin{verbatim}
family.glmm$c(args)
family.glmm$cp(args)
family.glmm$cpp(args)
\end{verbatim}

For the Bernoulli distribution, we calculate these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) with careful computer arithmetic as follows. We also use the \texttt{log1p} function in R for $c(\eta_i)$. 
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}
 For poisson, these values (\texttt{c}, \texttt{cp}, and \texttt{cpp}) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

@@

In the R function \texttt{glm}, the user can choose the link. The canonical link must be used in the \texttt{glmm} package so that we have an exponential family. The canonical link is included in \texttt{family.glmm}  in case the user does not know it already.

Also, I have a function to check that the data are valid given the family type. If \texttt{family.glmm} is \texttt{bernoulli.glmm}, the data should be 0s or 1s. If \texttt{family.glmm} is \texttt{poisson.glmm}, then the data should be nonnegative integers.  If the data does not pass the check, the check returns an error message.

\subsection{Redone in C}
Redone in C, I  have  separate functions for calculating $c(\eta)$, $c'(\eta)$, $c''(\eta)$: \texttt{cum3.c}, \texttt{cp3.c}, and \texttt{cpp3.c}. The inputs for each function are identical:  \texttt{eta} (an array of doubles), $\texttt{neta}$ (the length of \texttt{eta}), the type (to denote the family: $1$ indicates Bernoulli and $2$ indicates Poisson), and an array of doubles to contain the result. These are passed in as pointers. The functions  calculate  the cumulant function or one of its first two derivatives. Each function  contains a switch statement for the glmm family. The calculations for each of these functions have  been shown earlier in this section. This function is type void: rather than returning the cumulant or its derivatives, the pointers are changed to contain the results. This function is invoked by \texttt{el}, described in section \ref{sec:el}.

@@

\subsection{Redone in C}
The C function to calculate the value of the log data density and its two derivatives are called by reference. The following pointers are passed in: double \texttt{Y}, double \texttt{X}, int \texttt{nrowX}, int \texttt{ncolX}, double \texttt{eta}, int \texttt{family}, double \texttt{elval}, double \texttt{elgradient}, double \texttt{elhessian}. The pointers  \texttt{elval}, \texttt{elgradient} and \texttt{elhessian} are zeros before \texttt{el.C} is invoked. Invoking \texttt{el.C}  then places the calculatedvalue of the log data density and two derivatives into \texttt{elval}, \texttt{elgradient} and \texttt{elhessian}.\\

The function \texttt{el.C} calls the following C functions: \texttt{cum3.C} to calculate the cumulant given a value of $\eta$, \texttt{cp3.C} to calculate the derivative of the cumulant given a value of $\eta$, \texttt{cpp3.C} to calculate the hessian of the cumulant given a value of $\eta$, and functions to perform matrix multiplication.

\section{Random effect generation and calculations}
This section is focused on a vector of random effects $u$ with length $q$. We assume that $u \sim N(0,D)$ where $D$ depends on $\nu$. Section \ref{sec:getEk} details the  relationship between  $\nu$ and $D$ explicit. Section \ref{sec:genRand} describes the process of generating the random effects. Section \ref{sec:distRandsec} explains how to evaluate the distribution of the random effects and its first two derivatives.

%%getEk
\subsection{Constructing $D$ (\texttt{getEk})}\label{sec:getEk}
For this version of the package, we that assume that $D$ is diagonal.  Let $\nu$ be length $T$ with components $\nu_t$, and let $E_t,t=1,\ldots,T$ be diagonal matrices with indicators on the diagonal so that $\sum_{t=1}^T E_t = I$. That is, the diagonal entries  of $E_t$ indicate whether that random effect has $\nu_t$ as a variance component. Then $D=\sum_{t=1}^T\nu_t  E_t $.

If $q_t$ is the number of random effects associated with variance component $\nu_t$, then $q_t$ is also the number of nonzero entries in $E_t$ and $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model.  
The \texttt{model.matrix} part of \texttt{glmm} has made  \texttt{Z} into a list with $T$ design matrices (one for each distinct variance component).  In order to create $E_t$, we need to go through and count the  number of columns ($q_t$) for each design matrix $t$ in the list.
 Then $q=\sum_{t=1}^T q_t$ is the total number of random effects in the model, and $D$ will have $q$ rows (and columns). Thus, each $E_t$ will have $q$ rows and columns as well.  Then we know that $E_1$ has $q_1$ ones on the diagonal, followed by zeros to fill the rest of the diagonal.  $E_2$ has $q_1$ zeros, then $q_2$ ones, then zeros for the rest of the diagonal.

\subsection{Generating random effects (genRand)}\label{sec:genRand}
The model fitting function \texttt{glmm}  calls the function that generates the random effects and  stores them for future calculations. Let $s$ be a vector of length $q$ that represents the random effects on the standard normal scale. That is, $u= D^{1/2}s$. Let $\sigma$ be a vector of length $T$ with components $\sqrt{\nu_t}$.  The  PQL step of \texttt{glmm} has resulted in PQL estimates for $\beta$, $\sigma$ and  $s$ denoted by $\beta^*$, $s^*$ and $\sigma^*$. Then let
\begin{align}
A^*=\sum_{t=1}^T E_t \sigma^*_t
\end{align}\
and
\begin{align}
 D^* =A^*A^*.
\end{align}
Also, we can ``unstandardize'' our random effects:
\begin{align}
u^*&=A^*s^*.
\end{align}

Let $p_1,p_2,p_3$ be proportions such that  $p_1+p_2+p_3=1$. Let $D$ be determined by the current iteration of \texttt{trust}. The importance sampling distribution $\tilde{f}$ is the following mixture distribution:
\begin{align}
 \tilde{f}(u) = p_1  f(u \, | \, 0, D)+p_2  f(u \, | \, u^*, D^*)+p_3  f(u \, | \, u^*, (Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})
\end{align}

Therefore, the random effects are generated from the following distributions:
\begin{enumerate}
\item $N(0,I)$
\item $N(u^*,D^*)$
\item $N(u^*,(Z'  c''(X \beta^*+Zu^*) Z +(D^*)^{-1}   )^{-1})$.
\end{enumerate}

The first distribution is the multivariate standard normal, but the random effects drawn from this are scaled in the trust objective function so that their variance matrix $D$ is determined by the current iteration of trust.   The second distribution is centered at the PQL best guess of the random effect values $u^*$ and  has the PQL guess of the variance. The last distribution is centered at the PQL guess $u^*$ and has a variance based on the PQL penalized likelihood Hessian.  The idea of this last distribution is to generate random effects from a distribution whose Hessian matches that of the target distribution $f_\theta(y|u) f_\theta(u)$.

The random effects are generated by shifting and scaling a sample from a  standard normal distribution.  Suppose we want to generate a sample from a general normal distribution $N(\mu,\Sigma)$.  Let $u_k$ be a vector with length $q$ and let it denote one vector of generated random effects.  For each k, draw a sample of size $q$ from the standard normal distribution. Call this sample $\breve{u}_k$. Then shift and scale it:
\begin{align}
u_k = \breve{u}_k \Sigma^{1/2} + \mu.
\end{align}
so that is a sample from $N(\mu,\Sigma)$.
